The unphysicality of Hilbert spaces

### 2023-08-11 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Incorporating feedback

\begin{abstract}
We show that Hilbert spaces should not be considered the ``correct'' spaces to represent quantum states mathematically. We first prove that the requirements posited by complex inner product spaces are physically justified. We then show that completeness in the infinite-dimensional case requires the inclusion of states with infinite expectations, coordinate transformations that take finite expectations to infinite ones and vice-versa, and time evolutions that transform finite expectations to infinite ones in finite time. This makes Hilbert spaces physically unsound as they model a potential infinity as an actual infinity. We argue that Schwartz spaces may be better candidates, as, among other things, guarantee that the expectation of all polynomials of position and momentum are finite, guarantee solution to the moment problem and are the only space closed under Fourier transform. We suspect that at least some problems in quantum theory related to infinities may be ultimately caused by the wrong space being used.
\end{abstract}

@@

In an effort to find a precise physical justification, we realized that no such justification can exist as infinite-dimensional Hilbert spaces hopelessly contain unphysical elements. To put it simply, they must include states that have infinite expectation values for position, momentum, energy and so on. This means that any general statement in quantum theory must be valid also on these irrelevant unphysical cases, leaving out general statements that are only valid on the more constrained and better behaved physical ones. We therefore suspect that some of the mathematical difficulties encountered when trying to give a more precise mathematical foundation to quantum theory stem from this fundamental mismatch.

@@

Since we can physically prepare statistical ensembles, a physical theory must be able to characterize them. Mathematically, these form a convex set, where the convex combinations are the statistical mixtures.\footnote{There is a rich literature on generalized probabilistic theories (GPTs) in which people reconstruct quantum theory by imposing requirements on convex spaces.\cite{gpt_overview_2021}} In quantum mechanics, the geometry of the space of statistical ensembles is exactly defined by the complex inner product and vector space structures, which means properties of the inner product space can be reexpressed as properties of the space of mixtures. For example, we can characterize a superposition in the following way: a pure state $\psi$ is expressible as a superposition of other pure states $\{\phi_i\}$ if and only if there is a mixed state $\rho$ that can be equivalently expressed as a mixture of $\{\phi_i\}$ or as a mixture of $\psi$ and other pure states. That is, the ability to create a superposition is conceptually equivalent to the ability to prepare the same mixture in different ways, another feature particular to quantum mechanics. Additionally, the inner product can be shown to be equivalent to determining the entropy of all equal mixtures of pairs of states. These requirements are necessary in the sense that both classical and quantum mechanics need to keep track of how ensembles combine under mixing and what entropy each ensemble has. As such, this approach avoids the issues of interpretation.

@@

While we do need infinite-dimensional spaces to handle unbound quantities, such as the number of particles, or continuous quantities, such as position, it seems that Cauchy-completeness is not the proper characterization. While finding the proper one goes beyond the scope of this work, we want to show that alternatives are possible.

@@

\begin{itemize}
\item its elements are fully identified by the expectations - two Schwartz functions are the same if and only if all expectations are the same; quantum tomography on all polynomials fully reconstructs the state
\item it is an inner product space with the same norm as $L^2$
\item it is a dense subset of $L^2$ - any element of the Hilbert space can be approximated by a Schwartz function with an arbitrary level of precision
\item it is complete with respect to the expectation values - a sequence of functions for which all expectation values of all polynomials of position and momentum converge will converge in the Schwartz space
\item its closure over all Cauchy sequences recovers the Hilbert space - the only thing missing are the Cauchy sequences for which the limit of at least one expectation is not well defined
\item it is the only space closed under Fourier transform - the Fourier transform of a Schwartz function is a Schwartz function
\item it plays a fundamental role in the theory of distributions - objects like the $\delta$-function are mathematically defined on top of the Schwartz space
\end{itemize}
These properties make the Schwartz space a much more reasonable candidate to capture the physics. The convergence of expectation values of actual observables replaces convergence over vectors in an abstract space.\footnote{Compare this statement with the ``weak convergence'' of von Neumann algebras.} The point is that alternatives exist and should be fully explored.

@@

As for the inner product, we show that the Born rule can be understood as characterizing the entropy of equal mixtures over state pairs. This gives us a geometric understanding of the inner product at equal time instead of as a probability of transition, which brings it more in line with the use during superposition. It should also be noted that the von Neumann entropy of a mixture can be understood as the minimum Shannon entropy over all possible measurement contexts,\cite{nielsen_chuang_2010} which again shows the connection between the geometry of the convex space of statistical ensembles and the inner product.

@@

More concretely, a change of variable $X$ to $Y$ would be a unitary operator as it is just a change of representation. However, we can construct changes of variables such that the domain of $X$ is not exactly mapped to the domain of $Y$. Let $y=y(x)$ be our change of variable such that $y(0) = 0$. If $\psi(x)$ is a wavefunction expressed over $x$, and $\phi(y)$ represents the same state as a wavefunction over $y$, we have the following relationship:
\begin{equation}
\int_{0}^{y(x)} \phi^\dagger(y) \phi(y) dy = \int_{0}^{x} \psi^\dagger(x) \psi(x) dx
\end{equation}
To find a coordinate transformation that takes finite expectations to infinite expectations, we set the initial and final wavefunctions as follows:
\begin{align}
\psi(x) &= \sqrt{\frac{e^{-x^2}}{\sqrt{\pi}}} \\
\phi(y) &= \frac{1}{\sqrt{\pi(y^2 + 1)}}.
\end{align}
The first distribution is a Gaussian; therefore, all moments of the distribution, the expectation for $x^n$, are finite. The second goes to infinity like $\frac{1}{x^2}$, so the expectation will diverge for all moments above the first. We find
\begin{equation}
\begin{aligned}
\int_{0}^{y(x)} \frac{1}{\pi(\hat{y}^2 + 1)} d\hat{y} &= \int_{0}^{x} \frac{e^{-\hat{x}^2}}{\sqrt{\pi}} d\hat{x} \\
\frac{\tan^{-1}(y(x))}{\pi} &= \frac{\erf(x)}{2} \\
y(x) &= \tan \left(\frac{\pi}{2}\erf(x)\right). \\
\end{aligned}
\end{equation}
The graph of the above function increases so fast that, when plotted, it looks like it has a vertical asymptote. Nonetheless, it is an invertible function over the whole domain and will induce a unitary transformation on the Hilbert space.\footnote{Note that we could implement a similar coordinate transformation in classical mechanics, which would also change finite moments of a distribution into infinite ones. It may be that, like differentiable manifolds are constrained to only differentiable coordinate transformations, we need ``statistical manifolds'' that are constrained to coordinate transformations that preserve the finiteness of moments of distributions. This is a topic that requires further analysis.}

>>>>

\begin{abstract}
We show that Hilbert spaces should not be considered the ``correct'' spaces to represent quantum states mathematically. We first prove that the requirements posited by complex inner product spaces are physically justified. We then show that completeness in the infinite-dimensional case requires the inclusion of states with infinite expectations, coordinate transformations that take finite expectations to infinite ones and vice-versa, and time evolutions that transform finite expectations to infinite ones in finite time. This makes Hilbert spaces physically unsound as they model a potential infinity as an actual infinity. We suspect that at least some problems in quantum theory related to infinities may be ultimately caused by the wrong space being used. We strongly believe a better solution can be found, and we look at Schwartz spaces for inspiration, as, among other things, they guarantee that the expectation of all polynomials of position and momentum are finite, guarantee solution to the moment problem and are the only space closed under Fourier transform.
\end{abstract}

@@

In an effort to find a precise physical justification, we realized that no such justification can exist as infinite-dimensional Hilbert spaces hopelessly contain unphysical elements. To put it simply, they must include states that have infinite expectation values for position, momentum, energy and so on. This means that a truly general statement in quantum theory must be valid also on these irrelevant unphysical cases, with the consequence that either physical results are left out, or they are obscured in a complicated patchwork of theorems with different domains of validity. We therefore suspect that some of the mathematical difficulties encountered when trying to give a more precise mathematical foundation to quantum theory stem from this fundamental mismatch.

@@

Since we can physically prepare statistical ensembles, a physical theory must be able to characterize them. Mathematically, these form a convex set, where the convex combinations are the statistical mixtures.\footnote{There is a rich literature on generalized probabilistic theories (GPTs) in which people reconstruct quantum theory by imposing requirements on convex spaces.\cite{gpt_overview_2021}} In quantum mechanics, the geometry of the space of statistical ensembles is exactly defined by the complex inner product and vector space structures, which means properties of the inner product space can be reexpressed as properties of the space of mixtures. For example, we can characterize a superposition in the following way: a pure state $\psi$ is expressible as a superposition of other pure states $\{\phi_i\}$ if and only if there is a mixed state $\rho$ that can be equivalently expressed as a mixture of $\{\phi_i\}$ or as a mixture of $\psi$ and other pure states. That is, the ability to create a superposition is conceptually equivalent to the ability to prepare the same mixture in different ways, another feature particular to quantum mechanics. Additionally, the norm of the inner product, together with relative phases, can be shown to be equivalent to determining the entropy of all equal mixtures of pairs of states. These requirements are necessary in the sense that both classical and quantum mechanics need to keep track of how ensembles combine under mixing and what entropy each ensemble has. As such, this approach avoids the issues of interpretation.

@@

While we do need infinite-dimensional spaces to handle unbound quantities, such as the number of particles, or continuous quantities, such as position, it seems that Cauchy-completeness is not the proper characterization. While finding the proper one goes beyond the scope of this work, we want to show that fruitful lines of inquiry do exist.

@@

\begin{itemize}
\item its elements are fully identified by the expectations - two Schwartz functions are the same if and only if all expectations are the same; quantum tomography on all polynomials fully reconstructs the state
\item it is an inner product space with the same norm as $L^2$
\item it is a dense subset of $L^2$ - any element of the Hilbert space can be approximated by a Schwartz function with an arbitrary level of precision
\item it is complete with respect to the expectation values - a sequence of functions for which all expectation values of all polynomials of position and momentum converge will converge in the Schwartz space
\item its closure over all Cauchy sequences recovers the Hilbert space - the only thing missing are the Cauchy sequences for which the limit of at least one expectation is not well defined
\item it is the only space closed under Fourier transform - the Fourier transform of a Schwartz function is a Schwartz function
\item it plays a fundamental role in the theory of distributions - objects like the $\delta$-function are mathematically defined on top of the Schwartz space
\end{itemize}
These properties make the Schwartz space a much more reasonable candidate to capture the physics. The convergence of expectation values of actual observables replaces convergence over vectors in an abstract space.\footnote{Compare this statement with the ``weak convergence'' of von Neumann algebras.} To again be clear, we are not proposing this as a solution. We are merely pointing out that alternatives do exist and should be fully explored.\footnote{Note that $C^*$-algebras model observables and therefore do not solve the problem of modeling at hand.}

@@

As for the inner product, we show that the Born rule can be understood as characterizing the entropy of equal mixtures over state pairs. Note that the Born rule does not depend on the absolute phase, but it does depend on relative phases. If one expands a vector using superposition, in fact, the relative phase will introduce interference terms. Therefore recovering the Born rule means recovering the inner product up to total phases.

Associating the Born rule with entropy gives us a geometric understanding of the inner product at equal time instead of as a probability of transition, which brings it more in line with the use during superposition. It should also be noted that the von Neumann entropy of a mixture can be understood as the minimum Shannon entropy over all possible measurement contexts,\cite{nielsen_chuang_2010} which again shows the connection between the geometry of the convex space of statistical ensembles and the inner product.

@@

More concretely, a change of variable $x$ to $y$ induces a unitary operator as it is just a change of representation. The transformation rules for wavefunctions simply follow by the fact that $\psi^\dagger(x) \psi(x) = \rho(x)$ gives us a probability density over $x$. Therefore we must have
\begin{equation}
	\begin{aligned}
		\rho(x) dx &= \rho(y) dy \\
		\rho(y) &= \rho(x) \frac{dx}{dy} \\
		\psi^\dagger(y) \psi(y) &= \psi^\dagger(x) \psi(x) \frac{dx}{dy} \\
		\psi(y) &= \psi(x) \sqrt{\frac{dx}{dy}}.
	\end{aligned}
\end{equation}
This is a unitary transformation.

\begin{prop}
	Let $\mathcal{H} = L^2{\mathbb{R}}$ be the space of square integrable functions. Let $f : \mathbb{R} \to \mathbb{R}$ be an orientation-preserving diffeomorphism. Then the map $U : \mathcal{H} \to \mathcal{H}$ such that 
	$\phi(y) = U(\psi(x)) \mapsto \psi(x(y)) \sqrt{\left|\frac{df}{dx}\right|^{-1}}$ is a unitary transformation.
\end{prop}
\begin{proof}
	We have:
	\begin{equation}
		\begin{aligned}
			\int \phi^\dagger(y) \phi(y) dy &= \int U(\psi(x))^\dagger U(\psi(x)) dy \\
			 = \int \psi^\dagger(x(y)) &\sqrt{\left|\frac{df}{dx}\right|^{-1}}\psi(x(y)) \sqrt{\left|\frac{df}{dx}\right|^{-1}} dy \\
			 &= \int \psi^\dagger(x(y)) \psi(x(y)) \left|\frac{df}{dx}\right|^{-1} dy \\
			 &= \int \psi^\dagger(x) \psi(x) dx.
		\end{aligned}
	\end{equation}
	The map $U$ preserved the inner product and therefore is a unitary transformation.
\end{proof}

The issue is that the position operator $X$ and the position operator $Y$ are different operators, and the domain of $X$ will be in general different from the domain of $Y$. Let $y=y(x)$ be our change of variable such that $y(0) = 0$. If $\psi(x)$ is a wavefunction expressed over $x$, and $\phi(y)$ represents the same state as a wavefunction over $y$, we have the following relationship:
\begin{equation}
\int_{0}^{y(x)} \phi^\dagger(y) \phi(y) dy = \int_{0}^{x} \psi^\dagger(x) \psi(x) dx
\end{equation}
To find a coordinate transformation that takes finite expectations to infinite expectations, we set the initial and final wavefunctions as follows:
\begin{align}
\psi(x) &= \sqrt{\frac{e^{-x^2}}{\sqrt{\pi}}} \\
\phi(y) &= \frac{1}{\sqrt{\pi(y^2 + 1)}}.
\end{align}
The first distribution is a Gaussian; therefore, all moments of the distribution, the expectation for $x^n$, are finite. The second goes to infinity like $\frac{1}{x^2}$, so the expectation will diverge for all moments above the first. We find
\begin{equation}
\begin{aligned}
\int_{0}^{y(x)} \frac{1}{\pi(\hat{y}^2 + 1)} d\hat{y} &= \int_{0}^{x} \frac{e^{-\hat{x}^2}}{\sqrt{\pi}} d\hat{x} \\
\frac{\tan^{-1}(y(x))}{\pi} &= \frac{\erf(x)}{2} \\
y(x) &= \tan \left(\frac{\pi}{2}\erf(x)\right). \\
\end{aligned}
\end{equation}
The graph of the above function increases so fast that, when plotted, it looks like it has a vertical asymptote. Nonetheless, it is an invertible function over the whole domain and will induce a unitary transformation on the Hilbert space.\footnote{Note that we could implement a similar coordinate transformation in classical mechanics, which would also change finite moments of a distribution into infinite ones. It may be that, like differentiable manifolds are constrained to only differentiable coordinate transformations, we need ``statistical manifolds'' that are constrained to coordinate transformations that preserve the finiteness of moments of distributions. This is a topic that requires further analysis.}
